---
title: "Cyclistic - Data Process"
output: pdf_document
date: "2023-08-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidytext)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(tidyverse)

```

## Import and bind data into data frame 
```{r import_join_datas}

# use this code if data_all.RData not available

# data_all <- rbind(read.csv("dataclean/202207-clean.csv"), 
#               read.csv("dataclean/202208-clean.csv"), 
#               read.csv("dataclean/202209-clean.csv"),
#               read.csv("dataclean/202210-clean.csv"),
#               read.csv("dataclean/202211-clean.csv"),
#               read.csv("dataclean/202212-clean.csv"),
#               read.csv("dataclean/202301-clean.csv"),
#               read.csv("dataclean/202302-clean.csv"),
#               read.csv("dataclean/202303-clean.csv"),
#               read.csv("dataclean/202304-clean.csv"),
#               read.csv("dataclean/202305-clean.csv"),
#               read.csv("dataclean/202306-clean.csv"))

# load save environtment if data_all.RData available
load("data_all.RData")

```

```{r count_data}
head(data_all)
count(data_all)
```
## Format data datetime to correct it data types

format started_at and ended_at from char to datetime or S3: POSIXct
```{r format_datetime_data}

data_all$started_at <- as.POSIXct(data_all$started_at, format="%Y-%m-%d %H:%M:%S", tz="UTC")
data_all$ended_at <- as.POSIXct(data_all$ended_at, format="%Y-%m-%d %H:%M:%S", tz="UTC")

head(data_all)
```

## count time duration in each data and join it into table
```{r join_count_duration}

data_all$time_total <- data_all$ended_at - data_all$started_at

head(data_all)

```
## separate year, month, and day in each data and join it into table
```{r time_separate}

data_all <- data_all %>% 
  mutate(year = year(started_at), 
         month = month(started_at), 
         day = day(started_at))
  
head(data_all)
```


## save data into .RData files
```{r save_bind_join_data}

#save environtment data_all
# save(data_all, file="data_all.RData")

```

## Visualize data

### Visualize total ride of casual and member

#### a. Total ride in a year (July 2022 - June 2023)
```{r tot_year}
# count total ride for casual and member in bilion

ride_total <- data.frame(label=c("member", "casual"),
                         total=c(count(data_all[data_all$member_casual == "member", ])[1,1],
                                count(data_all[data_all$member_casual == "casual", ])[1,1]
                                         )
                         )

ride_total


```
```{r vis_tot_year}

# Visualize into barchart
ride_total %>% 
  ggplot(aes(x=label, y=total/1000, fill=label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(sprintf("%.0f", total/1000), "K")), color = "red") +
  ylab("Total bike in thousand") +
  xlab("Member_casual")

```

From visualize we know that member users is bike more often than casual users


#### b. Total ride monthly (July 2022 - June 2023)
```{r set_month_data}
# initialize data frame
ride_total_month <- data.frame(month = c("jul", "aug", "sep", "oct", "nov", "dec", "jan", "feb", 
                                         "march", "apr", "mei", "jun",
                                         "jul", "aug", "sep", "oct", "nov", "dec", "jan", "feb",
                                         "march", "apr", "mei", "jun"))
count_df <- data.frame(member_casual = c(NA), total = c(NA))

```

```{r get_month_data}
# set function to get month data using loop

count_ride <- function(data = data_all, mem_cas){
  #this function to take the ride data on data frame  
  
  if(mem_cas == "member") {
      count_data <-  count(data[data$member_casual == mem_cas & data$month == 7, ])[1,1]
      count_df <- data.frame(member_casual = c(mem_cas), total = c(count_data)) 
  }
  else {
    count_data <-  count(data[data$member_casual == mem_cas & data$month == 7, ])[1,1]
      count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, total = count_data)
  }
  for (i in 8:12) {
      count_data <-  count(data[data$member_casual == mem_cas & data$month == i, ])[1,1]
      count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, total = count_data)
  }
  for (i in 1:6) {
    count_data <-  count(data[data$member_casual == mem_cas & data$month == i, ])[1,1]
    count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, total = count_data)
  }
  
  return(count_df)
}

#get data ride
count_df <- count_df[nrow(count_df) + 1, ] <- count_ride(mem_cas = "member") 
count_df <- count_df[nrow(count_df) + 1, ] <- count_ride(mem_cas = "casual") 

```

```{r tot_ride_in_month}

# join into table
ride_total_month$member_casual <- count_df$member_casual
ride_total_month$total <-  as.numeric(count_df$total)

ride_total_month

```

```{r vis_tot_month}

# Visualize into barchart

ride_total_month %>% 
  ggplot(aes(fill=member_casual, y=total/1000, x=fct_inorder(month))) + 
  geom_bar(position="dodge", stat="identity") +
  ylab("Total bike in thousand") +
  xlab("Member_casual")

```
Form visualize we know that there are more bikers in middle of the year (in US this is spring time)

#### c. Total ride daily of all month (July 2022 - June 2023)
```{r set_day_data}
# initialize data frame

ride_total_days <- data.frame(dates = c(NA), member_casual = c(NA), total = c(NA))

```

```{r get_day_data}
# set function to get daily data using loop

count_ride <- function(data = data_all, mem_cas, count_df = NA){
  #this function to take the ride data on data frame  
  
  i = 1
  if(mem_cas == "member") {
    count_data <-  count(data[data$member_casual == mem_cas & data$day == i, ])[1,1]
    count_df <- data.frame(dates = c(i), member_casual = c(mem_cas), total = c(count_data)) 
  }
  else {
    count_data <-  count(data[data$member_casual == mem_cas & data$day == i, ])[1,1]
    count_df[nrow(count_df) + 1, ] <- c(dates = i, member_casual = mem_cas, total = count_data)
  }
  
  for (i in 2:31) {
    count_data <-  count(data[data$member_casual == mem_cas & data$day == i, ])[1,1]
    count_df[nrow(count_df) + 1, ] <- c(dates = i, member_casual = mem_cas, total = count_data)
    }
  return(count_df)
}

```

```{r tot_ride_in_days}

#get data ride
ride_total_days <- ride_total_days[nrow(ride_total_days) + 1, ] <- count_ride(mem_cas = "member") 
ride_total_days <- ride_total_days[nrow(ride_total_days) + 1, ] <- count_ride(mem_cas = "casual", count_df = ride_total_days) 

ride_total_days$total <-  as.numeric(ride_total_days$total)

ride_total_days

```

```{r vis_tot_day_in_month}

# Visualize into bar chart

ride_total_days %>% 
  ggplot(aes(fill=member_casual, y=total/1000, x=fct_inorder(dates))) + 
  geom_bar(position="dodge", stat="identity") +
  ylab("Total bike in thousand") +
  xlab("Member_casual")

```
From Visualize we know that bike data are distributed evenly for every dates 

### Visualize duration ride of casual and member
```{r separate_member_casual}
data_member <- data_all[data_all$member_casual == "member", ]
data_casual <- data_all[data_all$member_casual == "casual", ]
```

#### a. Average ride duration in a year (July 2022 - June 2023)
```{r dur_year}
# average ride duration for casual and member in bilion
 
ride_duration <- data.frame(member_casual = c("member", "casual"),
                            duration = c(
                              mean(data_member$time_total),
                              mean(data_casual$time_total)
                            ))

ride_duration

# print(sum(data_member$time_total))
```

```{r vis_dur_year}

# Visualize into barchart
ride_duration %>% 
  ggplot(aes(x=member_casual, y=as.numeric(duration)/60, fill=member_casual)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", as.numeric(duration)/60) ), color = "red") +
  ylab("Average bike duration in minutes") +
  xlab("Member_casual")

```
From visualize we know that casual users bike longer than member users

#### b. Average ride duration monthly (July 2022 - June 2023)
```{r set_month_dur_data}

# initialize data frame
ride_dur_month <- data.frame(month = c("jul", "aug", "sep", "oct", "nov", "dec", "jan", "feb", 
                                         "march", "apr", "mei", "jun",
                                         "jul", "aug", "sep", "oct", "nov", "dec", "jan", "feb",
                                         "march", "apr", "mei", "jun"))
avg_df <- data.frame(member_casual = c(NA), avg_dur = c(NA))

```

```{r get_month_dur_data}
# set function to get month data using loop

avg_duration <- function(data, mem_cas, count_df = NA ){
  #this function to take the average ride duration on data frame  
  
  if(mem_cas == "member") {
      count_data <-  mean(data$time_total[data$month == 7])
      count_df <- data.frame(member_casual = c(mem_cas), avg_dur = c(count_data))
  }
  else {
      count_data <-  mean(data$time_total[data$month == 7])
      count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, avg_dur = count_data)
  }
  for (i in 8:12) {
      count_data <-  mean(data$time_total[data$month == i])
      count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, avg_dur = count_data)
  }
  for (i in 1:6) {
      count_data <-  mean(data$time_total[data$month == i])
      count_df[nrow(count_df) + 1, ] <- c(member_casual = mem_cas, avg_dur = count_data)
  }
  return(count_df)
  
}

#get data duration
avg_df <- avg_df[nrow(avg_df) + 1, ] <- avg_duration(data = data_member, mem_cas = "member")
avg_df <- avg_df[nrow(avg_df) + 1, ] <- avg_duration(data = data_casual, mem_cas = "casual", count_df = avg_df) 

avg_df

```

```{r avg_dur_in_month}
# join into table
ride_dur_month$member_casual <- avg_df$member_casual
ride_dur_month$avg_dur <-  avg_df$avg_dur

ride_dur_month

```

```{r vis_avg_dur_month}

# Visualize into barchart

ride_dur_month %>%
  ggplot(aes(fill=member_casual, y=as.numeric(avg_dur)/60, x=fct_inorder(month))) +
  geom_bar(position="dodge", stat="identity") +
  ylab("Average bike duration in minutes") +
  xlab("Member_casual")

```
From visualize we know that peoples bike longer in middle of the year (in US this is a spring)

#### c. Average ride duration daily of all month (July 2022 - June 2023)
```{r set_day_dur_data}
# initialize data frame

avg_dur_days <- data.frame(dates = c(NA), member_casual = c(NA), avg_dur = c(NA))

```

```{r get_day_dur_data}
# set function to get daily data using loop

avg_duration <- function(data, mem_cas, count_df = NA ){
  #this function to take the average ride duration on data frame  
  i = 1
  if(mem_cas == "member") {
      count_data <-  mean(data$time_total[data$day == i])
      count_df <- data.frame(dates = c(i), member_casual = c(mem_cas), avg_dur = c(count_data))
  }
  else {
      count_data <-  mean(data$time_total[data$day == i])
      count_df[nrow(count_df) + 1, ] <- c(dates = i, member_casual = mem_cas, avg_dur = count_data)
  }
  
  for (i in 2:31) {
      count_data <-  mean(data$time_total[data$day == i])
      count_df[nrow(count_df) + 1, ] <- c(dates = i, member_casual = mem_cas, avg_dur = count_data)
  }
  return(count_df)
  
}

```

```{r avg_dur_in_days}

#get data duration
avg_dur_days <- avg_dur_days[nrow(avg_dur_days) + 1, ] <- avg_duration(data = data_member, mem_cas = "member")
avg_dur_days <- avg_dur_days[nrow(avg_dur_days) + 1, ] <- avg_duration(data = data_casual, mem_cas = "casual", count_df = avg_dur_days) 

avg_dur_days

```

```{r vis_avg_dur_day_in_month}

# Visualize into bar chart

avg_dur_days %>% 
  ggplot(aes(fill=member_casual, y=as.numeric(avg_dur)/60, x=fct_inorder(dates))) + 
  geom_bar(position="dodge", stat="identity") +
  ylab("Average bike duration in minutes") +
  xlab("Member_casual")

```

From Visualize we know that bike data are distributed evenly for every dates