---
title: "Cleaning Cylistic Data 2023-06"
output: pdf_document
date: "2023-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidytext)
library(dplyr)
library(ggplot2)

```

## Import data 
```{r import_datas}
data_01 <- read.csv(file="dataset/202306-divvy-tripdata.csv")
```

## Check data 01

Check the data type for each meta

```{r check_data_01}
str(data_01)
summary(data_01)
```
From meta check we know that data type of column "started_at" and "end_at" should be datetime


## Check duplicate data 01

```{r check_duplicate_01}
print(data_01[duplicated(data_01), ])

```

Duplicate data checking result : no data duplicate in data_01

## Remove duplicate data 

Remove Duplicate data result : No data to remove

## Check missing value data in character data type
```{r check_missing_value_01}
count(data_01[is.na(data_01$ride_id) | data_01$ride_id=="", ])
count(data_01[is.na(data_01$rideable_type) | data_01$rideable_type=="", ])
count(data_01[is.na(data_01$started_at) | data_01$started_at=="", ])
count(data_01[is.na(data_01$ended_at) | data_01$ended_at=="", ])
count(data_01[data_01$start_station_name=="", ])
count(data_01[data_01$start_station_id=="", ])
count(data_01[data_01$end_station_name=="", ])
count(data_01[data_01$end_station_id=="", ])
count(data_01[is.na(data_01$member_casual) | data_01$member_casual=="", ])
```

Missing value checking result : 

ride_id: [0]

rideable_type: [0]

started_at: [0]

ended_at: [0]

start_station_name: [116,259]

start_station_id: [116,259]

end_station_name: [124,050]

end_station_id:  [124,050]

member_casual: [0]

## Fill Missing value with NA

Missing value (empty data) in start_station_name, start_station_id, end_station_name, end_station_id will be filling with NA
```{r fill_missing_value_01}

data_01 <- replace(data_01, data_01 == "", NA)

```

Fill missing value result : empty data was replace with NA

## Check missing value data
```{r check_missing_value_coord_01}
count(data_01[is.na(data_01$start_lat) | data_01$start_lat=="", ])
count(data_01[is.na(data_01$start_lng) | data_01$start_lng=="", ])
count(data_01[is.na(data_01$end_lat) | data_01$end_lat=="", ])
count(data_01[is.na(data_01$end_lng) | data_01$end_lng=="", ])

```
Missing value checking result : 

start latitude and langitude : [0]

end latitude and langitude : [889]

## Remove Missing value with NA

Missing value in end_lat, end_lng will be delete by remove the row
```{r remove_missing_value_01}

# remove missing value data in this other data if there are also missing values 
# data_01 <- data_01[!is.na(data_01$rideable_type), ]
# data_01 <- data_01[!is.na(data_01$started_at), ]
# data_01 <- data_01[!is.na(data_01$ended_at), ]
# data_01 <- data_01[!is.na(data_01$member_casual), ]

data_01 <- data_01[!is.na(data_01$end_lat), ]
data_01 <- data_01[!is.na(data_01$end_lng), ]

count(data_01[is.na(data_01$end_lat) | data_01$end_lat=="", ])
count(data_01[is.na(data_01$end_lng) | data_01$end_lng=="", ])
```

Remove missing value result : Row with missing value data was removed 

## Check outliers in coordinate data
```{r check_outliers_01}
print(cat("start_lat : mean max min : ", 
    mean(data_01$start_lat), 
    max(data_01$start_lat), 
    min(data_01$start_lat)))
print(cat("start_lng : mean max min : ", 
    mean(data_01$start_lng), max(data_01$start_lng), min(data_01$start_lng)))
print(cat("end_lat : mean max min : ", 
    mean(data_01$end_lat), max(data_01$end_lat), min(data_01$end_lat)))
print(cat("end_lng : mean max min : ", 
    mean(data_01$end_lng), max(data_01$end_lng), min(data_01$end_lng)))

```
Outliers checking result : no outliers in coordinate data, max and min value for each data doesnt far from average value

## Remove useless column data

Acording to the bussines task, start_station_name and end_station_name will be remove
```{r remove_useless_data_01}

data_01 <- data_01[, -which(names(data_01) == "start_station_name")]
data_01 <- data_01[, -which(names(data_01) == "end_station_name")]

head(data_01)
str(data_01)

```

## Export clean data into csv
```{r export_01}

# write.csv(data_01, "dataclean/202306-clean.csv", row.names = FALSE)

```

